---
title: "Reproducible Research Course Project 1"
author: "bbreaden"
date: "March 9, 2018"
---


## Introduction

This is the Coursera Reproducible Research Course Project 1 created in html using knitr.

First some housekeeping (echo on, warning & message off).
```{r global_options, include=TRUE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

## 1. Loading and preprocessing the data

Load the zip file from the url supplied and unzip, then read data into the data frame 'dats'. Note: to save on data downloads and computing time, the code will only do this the first time (i.e. if files do not already exist).
```{r}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zips <- "dataset.zip"
file <- "activity.csv"
if(!file.exists(zips)){download.file(url,zips, mode = "wb")}
if(!file.exists(file)){unzip(zips, files = NULL, exdir=".")}
dats <- read.csv(file = file, header=TRUE)
str(dats)
```
Convert the dates (column 2) from factors to dates. (Happily they are in a tidy format, so this is simple.) 

## 2. Total steps per day

Calculate the total number of steps each day and store these in the array 'steps'. 
Use this calculate the mean and median of the steps per day, then display a frequency histogram of the steps per day, and the mean and median.
```{r}
dats$date <- as.Date(dats$date)
steps <- with(dats, tapply(steps, date, sum))
meansteps <- mean(steps, na.rm=TRUE)
medsteps <- median(steps, na.rm=TRUE)
paste("Mean of steps (ignoring missing values):", round(meansteps,2))
paste("Median of steps (ignoring missing values):", medsteps)
hist(steps, main="Frequency histogram of steps per day")
abline(v=meansteps, col="red", lty=1, lwd=2)
abline(v=medsteps, col="blue", lty=2, lwd=2)
```

In fact the mean (red) and median (blue) are so close together that they are indistinguishable on the histogram.

## 3. Daily pattern

Calculate number of steps per 5-minute interval, averaged across all days.
Store as array 'avesteps'. Make a line plot and calculate the interval with
the most steps on average. This interval is displayed in blue on the plot.
```{r}
means <- function(x){
        mean(x, na.rm=TRUE)
}
avesteps <- with(dats, tapply(steps, interval, means))
maxsteps <- which(avesteps==max(avesteps))
plot(x=names(avesteps),y=avesteps, type="l", lwd =2, xlab = "Interval (in seconds)", ylab = "Average steps")
abline(v=names(maxsteps), col="blue")
paste("Maximum average number of steps was", 
      round(avesteps[maxsteps],0), "during the 5-minute interval after the", 
      names(maxsteps), "minute mark.")
```

## 4. Calculate number of missing measurements.
```{r}
head(dats)
paste("In the original data frame there are", 
      sum(is.na(dats[,1])), "missing measurements out of",
      length(dats[,1]), "total measurements:",
      round(sum(is.na(dats[,1]))/length(dats[,1])*100,1),"%.")
```

## 5. Replace missing values and check.

The new dataset (I call it 'dats2') is identical to the original apart from these filled values:
```{r}
dats2 <- dats
for (i in 1:nrow(dats2)) {
    if (is.na(dats2$steps[i])) {
        dats2$steps[i] <- 
                avesteps[which(dats2$interval[i] == as.integer(names(avesteps)))]
    }
}
colSums(is.na(dats2))
head(dats2)
```

Plot another frequency histogram of the total steps per day, together with their mean (red) and median (blue).

```{r}
steps2 <- with(dats2, tapply(steps, date, sum))
hist(steps2, main="Frequency histogram of steps per day \n (missing data filled)")
abline(v=mean(steps2), col="red", lty=1, lwd=2)
abline(v=median(steps2), col="blue", lty=2, lwd=2)
paste("Mean steps before filling missing data was", round(meansteps,2), "and after filling was", round(mean(steps2),2),"This was the same because I used the mean values to fill the missing data.")
paste("Median steps before filling missing data was", medsteps, "and after filling was",median(steps2)," ...Slightly different.")
```

## 6. Differences between weekdays and weekends

First get the days of the week (in Mandarin because of my operating system...) and use them to create a new variable 'days' (either 'weekday' or 'weekend').
```{r}
dats2$days <- factor(format(dats2$date, "%A"))
table(dats2$days)
```

```{r}
levels(dats2$days) <- list(weekday = c("星期二","星期二","星期三","星期四","星期五"),
                           weekend = c("星期六","星期日"))
table(dats2$days)
```

Now create a panel plot the containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). (In the style in the Github ReadMe file.)

```{r}
avesteps3 <- aggregate(dats2$steps, 
                      list(interval = as.numeric(as.character(dats2$interval)), 
                           weekdays = dats2$days),
                      FUN = "mean")
names(avesteps3)[3] <- "meansteps"
library(lattice)
xyplot(avesteps3$meansteps ~ avesteps3$interval | avesteps3$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Steps")
```

ENDE